---
title: "Final Assignment Analysis of Spatiotemporal Data"
author: "Alexander Pilz"
date: "2023-01-30"
output:
  html_document:
    toc: yes
    theme: paper
  pdf_document:
    toc: yes
subtitle: Spatial point pattern analysis of wind turbines in the precincts of NRW
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

## Introduction

In July 2022, the German government formulated the "Act to Increase and Accelerate the Expansion of Onshore Wind Energy Facilities" which underscores the goal of increasing the use of renewable energy sources in Germany. The law stipulates that 1.8% of the area of the federal state of North Rhine-Westphalia (NRW) must be used for power generation from wind energy [8].
The power yields of such facilities are depended on a multitude of factors. One would argue that wind turbines should build where their potential energy yields are the highest. The planning and execution of construction efforts is the task of multiple administrative units. The federal state (Land NRW) and precincts (Bezirke) identify and disclose areas which are to be used for the power generation by means of wind energy in the federal state development plan (Landesentwicklungsplan) and the derivative regional plans (Regionalpläne). Some of the planning steps involved are subject to personal considerations. The counties (Kreise) are only responsible for the planning and subsequent construction of the actual wind parks. Thus the point pattern analysis is conducted on precinct level [9][10]. Instead of concerning itself with future planning practices the following point pattern analysis, conducted during the seminar Analysis of Spatio temporal data in 2022/23, tries to find and describe a possible relation between the point pattern formed by existing wind turbines in NRW and the environmental Variable of the specific energy power density. This in turn could shed some light on the driving factors which created the existing point pattern. Results such studies could be used to identify shortcomings of decision making processes and might be helpful for their optimization. 

## Data

The datasets employed during this point pattern analysis are all available openly and free of charge at the GEOportal.NRW.
The employed coordinate reference system of all datasets is ETRS89 / UTM zone 32N commonly used for official and cadastral data in NRW.
The analysis itself was carried out using the programming language R and the packages sf, raster, spatstat, plyr, maptools and their dependencies.

```{r importLibraries, echo=TRUE}
#import libraries
library(sf) 
library(raster)
library(spatstat)
library(plyr)
library(maptools)
```

The dataset containing polygonal features of the precincts of the federal state of NRW lists five precincts namely Arnsberg, Detmold, Köln, Düsseldorf and Münster (see Figure 1).

```{r summarizePrecincts, echo=TRUE, fig.height=8, fig.width=8}
#import precincts
precincts <- st_read("data/dvg2_EPSG25832_Shape/dvg2rbz_nw.shp")
#union precincts
precinctsUnion <- st_union(precincts)
```

```{r plotPrecincts, echo=FALSE, fig.height=8, fig.width=8}
plot(st_geometry(precincts), main = "Figure 1 - Precincts of NRW")
precinctsCentroids = st_coordinates(st_centroid(st_geometry(precincts)))
text(precinctsCentroids, labels = precincts$GN, col = "blue")
```

The point pattern to be examined represents the existing wind turbines in NRW.
The Shapefile contains 3797 point features with 12 attribute fields which contain information about the wind turbine like the diameter of its rotor, the year of its commissioning and the administrative unit in which they are situated (see Figure 2).

```{r summarizeTurbines, echo=TRUE}
#import wind turbines
windTurbines <- st_read("data/OpenEE-Windenergie_EPSG25832_Shape/Windbetrieb_Standorte.shp")
```

```{r plotTurbines, echo=FALSE, fig.height=8, fig.width=8}
plot(st_geometry(windTurbines), main = "Figure 2 - Wind turbines of NRW", pch=3, col="red" )
plot(precinctsUnion, add=TRUE)
```

The environmental variable to be used for this point pattern analysis, namely the specific energy power density (SEPD), was sourced as a raster file.
The raster has a resolution of 100x100 meters and covers all of NRW.
The data is available in different height levels above ground.
Here the data corresponding to a height of 100 meters is used.
The SEPD, ranging from 0 to 7, is a measure of the power of the wind flowing through an area.
It indicates how much power (in watts) is converted per square meter of rotor area.
The SEPD can be seen as an indicator for the suitability of a site for wind energy use (see Figure 3)[7].

```{r summarizeYields, echo=TRUE, fig.height=8, fig.width=8}
#import potential yields
potentialYields <- raster("data/Energieleistungsdichte-100m_EPSG25832_TIFF/en_100m_klas.tif")
#create vector of yields
yields <- c("1", "2", "3", "4", "5", "6", "7")
```

```{r plotYieldsI, echo=FALSE, fig.height=8, fig.width=8}
plot(potentialYields, main = "Figure 3 -  SEPD of NRW" )
```

## Methods

### Part I - Preliminary analysis and testing

In order to get an overview of the data the wind turbine counts and global intensities for each precinct are computed and visualized.
When computing the global intensities from the wind turbine point pattern the smoothing kernel bandwidth $sigma$ is chosen using a method described in [6].
Also Jones-Diggle improved edge corrections are applied.
Before analyzing the point pattern formed by the wind turbines in the precincts of NRW some preliminary tests are performed.
These encompass tests for complete spatial randomness.
The rejection of this is according to [3] "is a minimal prerequisite to any serious attempt to model an observed pattern" [3].
Therefore the empirical Functions are compared to a theoretical function which corresponds to a homogeneous poisson point process which is conforms to complete spatial randomness [2].

**Hypotheses Part I:**\
$H_{0} - I$: The wind turbines of NRW are distributed independently at random and uniformly in each precinct.\

$Ha_{1} - I$: The wind turbines of NRW exhibit a regular pattern and are not distributed independently at random and uniformly in each precinct.\

$Ha_{2} - I$: The wind turbines of NRW exhibit a clustered pattern and are not distributed independently at random and uniformly in each precinct.\

To confirm or reject these hypotheses the $G$-, $F$-, and $J$-functions were applied.
When computing the $G$-, $F$-, and $J$-functions the best edge correction method is chosen automatically.
The Monte-Carlo envelopes are computed simulating 99 point processes.
The $G$-function, also called the nearest-neighbour distance distribution function [2], "measures the distribution of the distances from an arbitrary event to its nearest event" [4].
In order assess the compatibility of the wind turbine point pattern with complete spatial randomness the empirical function $\hat{G}_{obs}$ is plotted against the theoretical function $(r)G_{theo}(r)$, thus the expectation under complete spatial randomness, and its Monte Carlo envelope indicated by $\hat{G}_{hi}(r)$ and $\hat{G}_{lo}(r)$ [4] for each precinct (see figures ).
Complete spatial randomness would be accepted for a given point process if the $\hat{G}_{obs}(r)$ function runs close to the $G_{theo}(r)$ function or within its Monte Carlo envelope.
A clustered pattern would be indicated if the empirical function $\hat{G}_{obs}(r)$ runs above the $G_{theo}(r)$ and its Monte Carlo envelope. Similarly a $\hat{G}_{obs}(r)$ function which runs below the $G_{theo}(r)$ and its Monte Carlo envelope would indicate a regular pattern [2][4].

The $F$-function, also called the empty-space function [2], "measures the distribution of all distances from an arbitrary point of the plane to its nearest event" [4].
Assessment of its results is done in a similar way as with the G-function by plotting empirical function $\hat{F}_{obs}(r)$ against the theoretical function $F_{theo}(r)$ and its Monte Carlo envelope indicated by $\hat{F}_{hi}(r)$ and $\hat{F}_{lo}(r)$ [4] (see figures ). One would accept complete spatial randomness for the $F$-function under the same conditions as for the $G$-function.
Regarding the indication of regular or clustered patterns, the reverse conditions as with the $G$-function must be applied [2][4].\

The $J$-function is a combination of the $G$- and $F$-function [2]. Assessment is again done by plotting the empirical function $\hat{J}_{obs}(r)$ against the theoretical function $J_{theo}(r)$ and its Monte Carlo envelope indicated by $\hat{J}_{hi}(r)$ and $\hat{J}_{lo}(r)$ (see figures ). The interpretation of the $J$-function can be more complex than with the $G$- and $F$-functions since it can also be used for "characterizing the interaction of points in terms of its type, strength and range". Clustered and regular patters are nonetheless indicated by the $J$-function as well where values below 1 would indicate the former and values above 1 the latter [1].

```{r subsetWindTurbines, echo=TRUE}
#create subset of wind turbines for each precinct
windTurbinesArnsberg <- st_intersection(windTurbines, precincts[precincts$GN == "Arnsberg",])
windTurbinesDetmold <- st_intersection(windTurbines, precincts[precincts$GN == "Detmold",])
windTurbinesKöln <- st_intersection(windTurbines, precincts[precincts$GN == "Köln",])
windTurbinesDüsseldorf <- st_intersection(windTurbines, precincts[precincts$GN == "Düsseldorf",])
windTurbinesMünster <- st_intersection(windTurbines, precincts[precincts$GN == "Münster",])

#create planar point patterns from wind turbine subsets using precincts as observation windows
pppArnsberg <- as.ppp(c(st_geometry(precincts[precincts$GN == "Arnsberg",]), st_geometry(windTurbinesArnsberg)))
pppDetmold <- as.ppp(c(st_geometry(precincts[precincts$GN == "Detmold",]), st_geometry(windTurbinesDetmold)))
pppKöln <- as.ppp(c(st_geometry(precincts[precincts$GN == "Köln",]), st_geometry(windTurbinesKöln)))
pppDüsseldorf <- as.ppp(c(st_geometry(precincts[precincts$GN == "Düsseldorf",]), st_geometry(windTurbinesDüsseldorf)))
pppMünster <- as.ppp(c(st_geometry(precincts[precincts$GN == "Münster",]), st_geometry(windTurbinesMünster)))

#create vector of precinct names and wind turbine counts
precintNames <- c("Arnsberg", "Detmold", "Köln", "Düsseldorf", "Münster")
windTurbineCounts <- c(pppArnsberg$n, pppDetmold$n, pppKöln$n, pppDüsseldorf$n, pppMünster$n)

#compute wind turbine density for each precinct
windTurbineDensities <- c(pppArnsberg$n/(as.numeric(st_area(precincts[precincts$GN == "Arnsberg",]))/1000000), 
                          pppDetmold$n/(as.numeric(st_area(precincts[precincts$GN == "Detmold",]))/1000000), 
                          pppKöln$n/(as.numeric(st_area(precincts[precincts$GN == "Köln",]))/1000000),
                          pppDüsseldorf$n/(as.numeric(st_area(precincts[precincts$GN == "Düsseldorf",]))/1000000),
                          pppMünster$n/(as.numeric(st_area(precincts[precincts$GN == "Münster",]))/1000000))

#compute precinct areas in km²
precintAreas <- c(as.numeric(st_area(precincts[precincts$GN == "Arnsberg",]))/1000000,
                  as.numeric(st_area(precincts[precincts$GN == "Detmold",]))/1000000,
                  as.numeric(st_area(precincts[precincts$GN == "Köln",]))/1000000,
                  as.numeric(st_area(precincts[precincts$GN == "Düsseldorf",]))/1000000,
                  as.numeric(st_area(precincts[precincts$GN == "Münster",]))/1000000)

#create dataframe with precinct names, areas, wind turbine counts and densities
precinctDf <- data.frame(precintNames, precintAreas, windTurbineCounts, windTurbineDensities)
colnames(precinctDf) <- c('Precinct','Area in km²','Turbine count', 'Turbine density')

#compute densities of wind turbines for each precinct
densityArnsberg <- density(pppArnsberg, sigma = bw.diggle, diggle = TRUE, edge = TRUE)
densityDetmold <- density(pppDetmold, sigma = bw.diggle, diggle = TRUE, edge = TRUE)
densityKöln <- density(pppKöln, sigma = bw.diggle, diggle = TRUE, edge = TRUE)
densityDüsseldorf <- density(pppDüsseldorf, sigma = bw.diggle, diggle = TRUE, edge = TRUE)
densityMünster <- density(pppMünster, sigma = bw.diggle, diggle = TRUE, edge = TRUE)
```

```{r plotWindTurbines, echo=FALSE, fig.show='hold', out.width='50%', fig.height=8, message=FALSE}
barplot(windTurbineCounts, main = "Figure 4 - Ammount of wind turbines per precinct", xlab = "Precinct", ylab = "Wind turbine count", names.arg = precintNames)
barplot(windTurbineDensities, main = "Figure 5 - Density of wind turbines per precinct", xlab = "Precinct", ylab = "Density per km²", names.arg = precintNames)
show(precinctDf)

par(mar=c(1, 1, 1, 1))
plot(st_geometry(precincts[precincts$GN == "Arnsberg",]), main = "Figure 6 - Wind turbines - Arnsberg" )
plot(st_geometry(windTurbinesArnsberg), pch = 3, col = 'red', add = TRUE)

plot(densityArnsberg, main = "Figure 7 - Wind turbine density - Arnsberg", box=FALSE)
plot(st_geometry(precincts[precincts$GN == "Arnsberg",]), add=TRUE)

plot(st_geometry(precincts[precincts$GN == "Detmold",]), main = "Figure 8 - Wind turbines - Detmold" )
plot(st_geometry(windTurbinesDetmold), pch = 3, col = 'red', add = TRUE)

plot(densityDetmold, main = "Figure 9 - Wind turbine density - Detmold", box=FALSE)
plot(st_geometry(precincts[precincts$GN == "Detmold",]), add=TRUE)

plot(st_geometry(precincts[precincts$GN == "Köln",]), main = "Figure 10 - Wind turbines - Köln" )
plot(st_geometry(windTurbinesKöln), pch = 3, col = 'red', add = TRUE)

plot(densityKöln, main = "Figure 11 - Wind turbine density - Köln", box=FALSE)
plot(st_geometry(precincts[precincts$GN == "Köln",]), add=TRUE)

plot(st_geometry(precincts[precincts$GN == "Düsseldorf",]), main = "Figure 12 - Wind turbines - Düsseldorf" )
plot(st_geometry(windTurbinesDüsseldorf), pch = 3, col = 'red', add = TRUE)

plot(densityDüsseldorf, main = "Figure 13 - Wind turbine density - Düsseldorf", box=FALSE)
plot(st_geometry(precincts[precincts$GN == "Düsseldorf",]), add=TRUE)

plot(st_geometry(precincts[precincts$GN == "Münster",]), main = "Figure 14 - Wind turbines of Münster precinct" )
plot(st_geometry(windTurbinesMünster), pch = 3, col = 'red', add = TRUE)

plot(densityMünster, main = "Figure 15 - Wind turbine density - Münster", box=FALSE)
plot(st_geometry(precincts[precincts$GN == "Münster",]), add=TRUE)
```

```{r gFunction, echo=FALSE, message=FALSE, warning=FALSE}
#compute G-, F- and J-function for precinct Arnsberg
gEnvArnsberg <- envelope(pppArnsberg, fun = Gest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
fEnvArnsberg <- envelope(pppArnsberg, fun = Fest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
jEnvArnsberg <- envelope(pppArnsberg, fun = Jest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)

#compute G-, F- and J-function for precinct Detmold
gEnvDetmold <- envelope(pppDetmold, fun = Gest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
fEnvDetmold <- envelope(pppDetmold, fun = Fest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
jEnvDetmold <- envelope(pppDetmold, fun = Jest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)

#compute G-, F- and J-function for precinct Arnsberg
gEnvKöln <- envelope(pppKöln, fun = Gest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
fEnvKöln <- envelope(pppKöln, fun = Fest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
jEnvKöln <- envelope(pppKöln, fun = Jest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)

#compute G-, F- and J-function for precinct Arnsberg
gEnvDüsseldorf <- envelope(pppDüsseldorf, fun = Gest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
fEnvDüsseldorf <- envelope(pppDüsseldorf, fun = Fest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
jEnvDüsseldorf <- envelope(pppDüsseldorf, fun = Jest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)

#compute G-, F- and J-function for precinct Arnsberg
gEnvMünster <- envelope(pppMünster, fun = Gest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
fEnvMünster <- envelope(pppMünster, fun = Fest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
jEnvMünster <- envelope(pppMünster, fun = Jest, nsim = 99, clipdata = TRUE, correction = "best", verbose = FALSE)
```

```{r gFunctionPlot, echo=FALSE, fig.show='hold', out.width='50%', fig.height=8, message=FALSE, warning=FALSE}
#plot G-function for precincts
plot(gEnvArnsberg, main = "Figure 16 - G-function - Arnsberg" )
plot(gEnvDetmold, main = "Figure 17 - G-function - Detmold")
plot(gEnvKöln, main = "Figure 18 - G-function - Köln")
plot(gEnvDüsseldorf, main = "Figure 19 - G-function - Düsseldorf")
plot(gEnvMünster, main = "Figure 20 - G-function - Münster")
```

```{r fFunctionPlot, echo=FALSE, fig.show='hold', out.width='50%', fig.height=8, message=FALSE, warning=FALSE}
#plot F-function for precincts
plot(fEnvArnsberg, main = "Figure 21 - F-function - Arnsberg")
plot(fEnvDetmold, main = "Figure 22 - F-function - Detmold")
plot(fEnvKöln, main = "Figure 23 - F-function - Köln")
plot(fEnvDüsseldorf, main = "Figure 24 - F-function - Düsseldorf")
plot(fEnvMünster, main = "Figure 25 - F-function - Münster")
```

```{r jFunctionPlot, echo=FALSE, fig.height=8, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
#plot J-function for precincts
plot(jEnvArnsberg, main = "Figure 26 - J-function - Arnsberg", ylim=c(0,12))
plot(jEnvDetmold, main = "Figure 27 - J-function - Detmold", ylim=c(0,12))
plot(jEnvKöln, main = "Figure 28 - J-function - Köln", ylim=c(0,12))
plot(jEnvDüsseldorf, main = "Figure 29 - J-function - Düsseldorf", ylim=c(0,12))
plot(jEnvMünster, main = "Figure 30 - J-function - Münster", ylim=c(0,12))
```

### Part II - Local intensity

A firsts step towards finding and understanding a possible relationship between a point pattern and corresponding values of an underlying variable could be a local density analysis using the SEPD as a tessellated covariate.
The densities are computed for the subareas formed by the pixels of equal value (see figures ).
Such an approach can be used if it is believed that the point pattern process is driven by the values of the underlying tessellated surface. Here the assumption is that areas with high potential power yields which are indicated by high SEPD values should exhibit higher wind turbine densities.

**Hypotheses Part II:**\
$H_{0} - II$: There is no linear correlation between local wind turbine densities and the SEPD values.\

$Ha_{1} - II$: There is a positive linear correlation between local wind turbine densities and the SEPD values.\

$Ha_{2} - II$: There is a negative linear correlation between local wind turbine densities and the SEPD values.\

A p-value of a Shapiro-Wilk test greater than 0.05 would imply that the distribution of the local wind turbine intensities are not significantly different from normal distribution. Since this is not the case the Kendall-tau rank correlation is used since it tends to be more robust against outliers than the Spearman rank correlation [5]. The local wind tubine intensities are ranked before computing the correlation coefficients. The correlation coefficient ranges form -1 to 1 where a positive value would indicate that high SEPD are accompanied by high wind turbine intensities while a negative value would indicate the opposite. The significance of a correlation can assessed by he accompanying p-value which should be smaller than 0.05.

```{r cropYields, echo=FALSE, fig.show='hold', out.width='50%', message=FALSE, warning=FALSE}
#crop  SEPD to extend of the precincts
potentialYieldsArnsbergCropped <- crop(potentialYields, precincts[precincts$GN == "Arnsberg",])
potentialYieldsArnsbergMasked <- mask(potentialYieldsArnsbergCropped, precincts[precincts$GN == "Arnsberg",])

potentialYieldsDetmoldCropped <- crop(potentialYields, precincts[precincts$GN == "Detmold",])
potentialYieldsDetmoldMasked <- mask(potentialYieldsDetmoldCropped, precincts[precincts$GN == "Detmold",])

potentialYieldsKölnCropped <- crop(potentialYields, precincts[precincts$GN == "Köln",])
potentialYieldsKölnMasked <- mask(potentialYieldsKölnCropped, precincts[precincts$GN == "Köln",])

potentialYieldsDüsseldorfCropped <- crop(potentialYields, precincts[precincts$GN == "Düsseldorf",])
potentialYieldsDüsseldorfMasked <- mask(potentialYieldsDüsseldorfCropped, precincts[precincts$GN == "Düsseldorf",])

potentialYieldsMünsterCropped <- crop(potentialYields, precincts[precincts$GN == "Münster",])
potentialYieldsMünsterMasked <- mask(potentialYieldsMünsterCropped, precincts[precincts$GN == "Münster",])
```

```{r localDensity, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
#create vector of breaks
breaks <- c(1,2,3,4,5,6,7)

#compute local intensity using the  SEPD as tessellated covariate
tessellationArnsberg <- tess(image=potentialYieldsArnsbergMasked)  
localDensityArnsberg <- quadratcount(pppArnsberg, tess=tessellationArnsberg)
localDensityArnsberg.dens <- intensity(localDensityArnsberg) 

tessellationDetmold <- tess(image=potentialYieldsDetmoldMasked)  
localDensityDetmold <- quadratcount(pppDetmold, tess=tessellationDetmold)  
localDensityDetmold.dens <- intensity(localDensityDetmold) 

tessellationKöln <- tess(image=potentialYieldsKölnMasked)  
localDensityKöln <- quadratcount(pppKöln, tess=tessellationKöln) 
localDensityKöln.dens <- intensity(localDensityKöln) 

tessellationDüsseldorf <- tess(image=potentialYieldsDüsseldorfMasked)  
localDensityDüsseldorf <- quadratcount(pppDüsseldorf, tess=tessellationDüsseldorf)  
localDensityDüsseldorf.dens <- intensity(localDensityDüsseldorf) 

tessellationMünster <- tess(image=potentialYieldsMünsterMasked)  
localDensityMünster <- quadratcount(pppMünster, tess=tessellationMünster)  
localDensityMünster.dens <- intensity(localDensityMünster) 

#compute correlation between  SEPD and (ranked) wind turbine intensity
localDensityArnsberg.df <- data.frame(localDensityArnsberg.dens)
localDensityArnsberg.df$densityRank <- rank(localDensityArnsberg.df$Freq, ties.method= "max")
localDensityArnsberg.cor <-cor.test(as.numeric(localDensityArnsberg.df$tile), localDensityArnsberg.df$densityRank,  method = "kendall") 

localDensityDetmold.df <- data.frame(localDensityDetmold.dens)
localDensityDetmold.df$densityRank <- rank(localDensityDetmold.df$Freq, ties.method= "max")
localDensityDetmold.cor <-cor.test(as.numeric(localDensityDetmold.df$tile), localDensityDetmold.df$densityRank,  method = "kendall") 

localDensityKöln.df <- data.frame(localDensityKöln.dens)
localDensityKöln.df$densityRank <- rank(localDensityKöln.df$Freq, ties.method= "max")
localDensityKöln.cor <-cor.test(as.numeric(localDensityKöln.df$tile), localDensityKöln.df$densityRank,  method = "kendall") 

localDensityDüsseldorf.df <- data.frame(localDensityDüsseldorf.dens)
localDensityDüsseldorf.df$densityRank <- rank(localDensityDüsseldorf.df$Freq, ties.method= "max")
localDensityDüsseldorf.cor <-cor.test(as.numeric(localDensityDüsseldorf.df$tile), localDensityDüsseldorf.df$densityRank,  method = "kendall") 

localDensityMünster.df <- data.frame(localDensityMünster.dens)
localDensityMünster.df$densityRank <- rank(localDensityMünster.df$Freq, ties.method= "max")
localDensityMünster.cor <-cor.test(as.numeric(localDensityMünster.df$tile), localDensityMünster.df$densityRank,  method = "kendall") 
```

```{r plotYields, echo=FALSE, fig.height=8, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
plot(potentialYieldsArnsbergMasked, main = "Figure 31 - SPED & wind turbines - Arnsberg", zlim = c(1, 7))
plot(st_geometry(windTurbinesArnsberg), pch = 3, col = 'black', add = TRUE)
barplot(localDensityArnsberg.dens, main = "Figure 32 - Local wind turbine densities - Arnsberg", xlab = "SEPD", ylab = "Local wind turbine density")

plot(potentialYieldsDetmoldMasked, main = "Figure 33 - SPED & wind turbines - Detmold", zlim = c(1, 7))
plot(st_geometry(windTurbinesDetmold), pch = 3, col = 'black', add = TRUE)
barplot(localDensityDetmold.dens, main = "Figure 34 - Local wind turbine densities - Detmold", xlab = "SEPD", ylab = "Local wind turbine density")

plot(potentialYieldsKölnMasked, main = "Figure 35 - SPED & wind turbines - Köln", zlim = c(1, 7))
plot(st_geometry(windTurbinesKöln), pch = 3, col = 'black', add = TRUE)
barplot(localDensityKöln.dens, main = "Figure 36 - Local wind turbine densities - Köln", xlab = "SEPD", ylab = "Local wind turbine density")

plot(potentialYieldsDüsseldorfMasked, main = "Figure 37 - SPED & wind turbines - Düsseldorf", zlim = c(1, 7))
plot(st_geometry(windTurbinesDüsseldorf), pch = 3, col = 'black', add = TRUE)
barplot(localDensityDüsseldorf.dens, main = "Figure 38 - Local wind turbine densities - Düsseldorf", xlab = "SEPD", ylab = "Local wind turbine density")

plot(potentialYieldsMünsterMasked, main = "Figure 39 - SPED & wind turbines - Münster", zlim = c(1, 7))
plot(st_geometry(windTurbinesMünster), pch = 3, col = 'black', add = TRUE)
barplot(localDensityMünster.dens, main = "Figure 40 - Local wind turbine densities - Münster", xlab = "SEPD", ylab = "Local wind turbine density")
```

```{r plotCorrelations, echo=FALSE, fig.show='hold', out.width='50%', fig.height=8, message=FALSE, warning=FALSE}
localDensityArnsberg.cor
localDensityDetmold.cor
localDensityKöln.cor
localDensityDüsseldorf.cor
localDensityMünster.cor
```

### Part III - Intensity as function of covariate

Intensities of a point processes can be modeled using Poisson point processes.
Additionally a covariate can be introduced if the assumption is made that the point process is driven or at least influenced by the covariate [2]. To evaluate the quality of the resulting model a Poisson point process model is fitted without an additional covariate.

**Hypotheses Part III:**\
$H_{0} - III$: The wind turbine point pattern is not depended on the SEPD.\

$Ha - III$: The wind turbine point pattern is depended on the SEPD.\

The comparison between the models is carried out using the likelihood ratio test.
The p-value ($Pr(>Chi)$) of such a test is interpreted as the probability of being wrong in rejecting the $H_{0}$ [2].

```{r intensityFunction, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
covariantArnsberg <- as.im.RasterLayer(potentialYieldsArnsbergMasked)
ppmArnsbergCov <- ppm(pppArnsberg ~ covariantArnsberg, data=c(tessellationArnsberg))
ppmArnsberg <- ppm(pppArnsberg ~ 1)
ppmArnsbergPred <- predict(ppmArnsbergCov)

covariantDetmold <- as.im.RasterLayer(potentialYieldsDetmoldMasked)
ppmDetmoldCov <- ppm(pppDetmold ~ covariantDetmold, data=c(tessellationDetmold))
ppmDetmold <-ppm(pppDetmold ~ 1)
ppmDetmoldPred <- predict(ppmDetmoldCov)

covariantKöln <- as.im.RasterLayer(potentialYieldsKölnMasked)
ppmKölnCov <- ppm(pppKöln ~ covariantKöln, data=c(tessellationKöln))
ppmKöln <- ppm(pppKöln ~ 1)
ppmKölnPred <- predict(ppmKölnCov)

covariantDüsseldorf <- as.im.RasterLayer(potentialYieldsDüsseldorfMasked, )
ppmDüsseldorfCov <- ppm(pppDüsseldorf ~ covariantDüsseldorf, data=c(tessellationDüsseldorf))
ppmDüsseldorf <- ppm(pppDüsseldorf ~ 1)
ppmDüsseldorfPred <- predict(ppmDüsseldorfCov)

covariantMünster <- as.im.RasterLayer(potentialYieldsMünsterMasked)
ppmMünsterCov <- ppm(pppMünster ~ covariantMünster, data=c(tessellationMünster))
ppmMünster <- ppm(pppMünster ~ 1)
ppmMünsterPred <- predict(ppmMünsterCov)
```

```{r plotIntensityFunction, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
plot(effectfun(ppmArnsbergCov, "covariantArnsberg", se.fit=TRUE), main="Figure 41 - Intensity function using SEPD as covariate - Arnsberg", las=1, xlab = "SEPD")
plot(ppmArnsbergPred, box=FALSE, main="Figure 42 - Modeled wind turbine intensity - Arnsberg")

plot(effectfun(ppmDetmoldCov, "covariantDetmold", se.fit=TRUE), main="Figure 43 - Intensity function using SEPD as covariate - Detmold", las=1, xlab = "SEPD")
plot(ppmDetmoldPred, box=FALSE, main="Figure 44 - Modeled wind turbine intensity - Detmold")

plot(effectfun(ppmKölnCov, "covariantKöln", se.fit=TRUE), main="Figure 45 - Intensity function using SEPD as covariate - Köln", las=1, xlab = "SEPD")
plot(ppmKölnPred, box=FALSE, main="Figure 46 - Modeled wind turbine intensity -  Köln")

plot(effectfun(ppmDüsseldorfCov, "covariantDüsseldorf", se.fit=TRUE), main="Figure 47 - Intensity function using SEPD as covariate - Düsseldorf", las=1, xlab = "SEPD")
plot(ppmDüsseldorfPred, box=FALSE, main="Figure 48 - Modeled wind turbine intensity -  Düsseldorf")

plot(effectfun(ppmMünsterCov, "covariantMünster", se.fit=TRUE), main="Figure 49 - Intensity function using SEPD as covariate - Münster", las=1, xlab = "SEPD")
plot(ppmMünsterPred, box=FALSE, main="Figure 50 - Modeled wind turbine intensity - Münster")
```

```{r testFunction, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, out.width='50%'}
anova(ppmArnsberg, ppmArnsbergCov, test="LR")
anova(ppmDetmold, ppmDetmoldCov, test="LR")
anova(ppmKöln, ppmKölnCov, test="LR")
anova(ppmDüsseldorf, ppmDüsseldorfCov, test="LR")
anova(ppmMünster, ppmMünsterCov, test="LR")
```

## Results

### Part I

The highest wind turbine counts are found in the precincst of Münster and Detmold with 1067 and 1002 wind turbines respectively.
The precincts of Köln and Arnsberg share similar wind turbine counts, 678 for the former and 633 for the latter.
The lowest turbine counts is obsersved in the Düsseldorf precinct.
The wind turbine densities per km² follow a similar pattern where the precincts of Detmold and Köln exhibit values around 0.15 wind turbines per km².
Here the Köln precinct exhibits a slightly higher wind turbine densities per km², 0.09, as the precincts of Arnsberg and Düsseldorf which share a wind turbine densities per km² 0f 0.07 (See figures 4, 5).
A purely visual inspection of the spatial distributions and densities of the wind turbines in the precincts of NRW suggests that the underlying point process might be clustered (See Figures 6 to 16).
This is confirmed by the $G$-, $F$-, and $J$-functions.
The $\hat{G}_{obs}(r)$ functions lie above the lower simulation envelope of the ${G}_{theo}(r)$ functions for most of their range (See figures 16-20).
The opposite can be observed with the the $\hat{F}_{obs}(r)$ functions.
These lie below the respective simulation envelope of ${F}_{theo}(r)$ function for the major part of their range (See figures 21-25).
The $\hat{J}_{obs}(r)$ functions exhibit values \< 1 for the majority of their range thus Conforming the results from the $G$- and $F$-functions (See figures 26-30).
Based on the results of the $G$-, $F$-, and $J$-functions the $H_{0} - I$ is rejected.
Based on the properties described in Methods Part I $Ha_{2} - I$ is accepted thus reaching the conclusion that the wind turbine point process in NRW exhibits a clustered pattern.

### Part II

A purely visual inspection of local intensity analysis hints towards a possible positive correlation between the local wind turbine intensity and the SEPD.
It has to mentioned that the wind turbine densities extreme of the SEPD mostly yielded local wind turbine intensities of 0.
The Kendalls-tau correlation coefficients, with the exception of the Detmold precinct, do not yield a significant correlation between the local wind turbine intensity and the SEPD.
The computed values for Kendalls-tau are nonetheless positive for all precincts.
It is of note that the values for Kendalls-tau are significant and close to 1  when the aforementioned extreme SEPD values are omitted during computation.
The $H_{0} - II$ can only be rejected for the precinct of Detmold due to the otherwise unsatisfactory p-values which leads to the conclusion that there is no significant linear correlation between the local wind turbine intensity and the SEPD in the precincts of Arnsberg, Köln, Düsseldorf and Münster. For the precinct of Detmold $Ha_{1} - II$ is accepted.

### Part III

Despite of the unsatisfactory levels of significance it was attempted to model the wind turbine density as a function of the SPED.
The resulting Poisson pint process using the SPED as a covariant was compared to another Poisson point process which was fitted to the wind turbine point pattern only. The comparison was carried out using likelihood ratio tests.
These yielded very low p-values ($Pr(>Chi)$). This leads to to the rejection of $H_{0} - III$ and the acceptance of $H_a - III$.
This however is not an indication that the SEPD is the only variable influencing the wind turbine point process. 

## Discussion

The presented analysis could prove that the point process formed by the wind turbines in the precincts of NRW is clustered.
This seems not surprising given that wind turbines tend to aggregated together in wind parks and are rarely build individually. The analysis of the local wind turbine intensities and their correlation with the underlying SEPD values could disprove the existence of a significant linear correlation between them in all precincts of NRW.
Further studies could employ different measures and methods to identify other a possible correlations between the two since positive though insignificant correlation coefficients could be seen as an indication that some relation between the two exists. 
The fitting of Poisson point processes to the wind turbine point pattern and the SEPD as a covariate and their subsequent comparison could prove that the wind turbine point pattern is depended on the SEPD though the degree of this dependence could not be determined although it seems highly unlikely that the SEPD is the only factor influencing the disclosure of areas which are to be used for the power generation by means of wind energy by the planning authorities.
Future attempts at modelling the underlying Poisson point processes of the wind turbine point pattern should therefore take additional Variables like regulatory constraints like the distance to residential areas, environmental variables like the SPED in other available height levels or the relief into account. It could also be worth while to conduct comparative studies with different processes for modelling the wind turbine point process like Cox processes [3].
Based on the described methods and results it could be argued that the specific energy power density might be seen as a factor which influenced the decisions which led to he formation of the wind turbine point pattern in the precincts of NRW. Further, more in-depth studies are nonetheless strongly recommended.

## Sources

[1] J.Illian, A. Penttinen, H. Stoyan and D. Stoyan, Statistical Analysis and Modelling of Spatial Point Patterns, 1st edition Chichester: John Wiley & Sons Ltd., 2008

[2] A. Baddeley, E. Rubak, R. Turner, Spatial Point Patterns Methodology and Applications with R, 1st edition Boca Raton: Taylor & Francis Group, LLC, 2016

[3] P. J. Diggle, Statistical Analysis of Spatial and Spatio-Temporal Point Patterns, 3rd edition Boca Raton: Taylor & Francis Group, LLC, 2014

[4] R.
S. Bivand, E. Pebesma, V. Gómez-Rubio, Applied Spatial Data Analysis with R, 2nd edition New York: Springer Science+Business Media 2013

[5] C.
Reimann, P. Filzmoser, R. G. Garrett, R. Dutter, Statistical Data Analysis Explained Applied Environmental Statistics with R, 1st edition Chichester: John Wiley & Sons Ltd., 2008

[6] P.J.
Diggle, Statistical analysis of spatial point patterns, 2nd edition Dortmund: Arnold, 2003

[7] Landesamt für Natur, Umwelt und Verbraucherschutz Nordrhein-Westfalen (LANUV), Potenzialstudie Windenergie NRW LANUV-Fachbericht 124 [Online], Available at: <https://www.lanuv.nrw.de/fileadmin/lanuvpubl/3_fachberichte/Potenzialstudie-Windenergie-NRW.pdf> (Last accessed on: 03. March 2023)

[8] Bundestag der Bundesrepublik Deustchland, Gesetz zur Erhöhung und Beschleunigung des Ausbaus von Windenergieanlagen an Land [Online], Available at: <https://www.bgbl.de/xaver/bgbl/start.xav#__bgbl__%2F%2F*%5B%40attr_id%3D%27bgbl122s1353.pdf%27%5D__1678185292530> (Last accessed on: 03. March 2023)

[9] Bundestag der Bundesrepublik Deustchland, Raumordnungsgesetz [Online], Available at: <https://www.gesetze-im-internet.de/rog_2008/ROG.pdf> (Last accessed on: 03. March 2023)

[10] Landesregierung Nordrhein-Westfalen, Landesplanungsgesetz Nordrhein-Westfalen [Online], Available at: <https://recht.nrw.de/lmi/owa/br_text_anzeigen?v_id=920070925160557909> (Last accessed on: 03. March 2023)
